---
title: "Reanalysis of the effect of Mst-1 overexpression in young and old mice"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/mst1_reanalysis

# Introduction

```{r,packages}

suppressPackageStartupMessages({
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("mitch")
    library("eulerr")
    library("kableExtra")
    library("beeswarm")
    library(ggplot2)
})

```

## Import DE data


```{r,importdata}

you <- read.table("Mst1TG_young_deseq.tsv",header=TRUE,sep="\t")

old <- read.table("Mst1TG_old_deseq.tsv",header=TRUE,sep="\t")

you_up <- rownames(subset(you,log2FoldChange > 0 & padj < 0.05))
you_dn <- rownames(subset(you,log2FoldChange < 0 & padj < 0.05))

old_up <- rownames(subset(old,log2FoldChange > 0 & padj < 0.05))
old_dn <- rownames(subset(old,log2FoldChange < 0 & padj < 0.05))

```


## Euler diagram

```{r,euler1}

par(mar=c(2,2,2,2))

v0 <- list("young up"=you_up, 
  "young dn"=you_dn,
  "old up"=old_up,
  "old dn"=old_dn)

plot(euler(v0),quantities = TRUE, edges = "gray", main="effect of Mst-1 overexpression")

```

## Mitch

First fetch genesets.

```{r,geneset1}

if  (!file.exists("mouse_msigdb_reactome_2022-02-16.gmt")) {
  download.file("http://ziemann-lab.net/public/msigdb_mouse/mouse_msigdb_reactome_2022-02-16.gmt",
    destfile="mouse_msigdb_reactome_2022-02-16.gmt")
}
genesets <- gmt_import("mouse_msigdb_reactome_2022-02-16.gmt")
names(genesets) <- gsub("REACTOME_","",names(genesets))
names(genesets) <- gsub("_"," ",names(genesets))

```

Now set up the gene table `gt`.

```{r,gt}

gt <- as.data.frame(unique(rownames(old) ,rownames(you)))
gt$gene <- sapply(strsplit(gt[,1]," "),"[[",2)

```

Now execute mitch analysis.

```{r,mitch1}

m1 <- mitch_import(list("young"=you,"old"=old), DEtype="deseq2",geneTable=gt)

cor(m1,method="pearson")
cor(m1,method="spearman")

mres1 <- mitch_calc(m1, genesets, priority="effect")

cor(mres1$enrichment_result$s.young, mres1$enrichment_result$s.old,method="pearson")
cor(mres1$enrichment_result$s.young, mres1$enrichment_result$s.old,method="spearman")


head(mres1$enrichment_result,20) %>% 
  kbl(caption = "Top gene pathway differences caused by Mst-1 overexpression") %>%
  kable_paper("hover", full_width = F)

```

Scatterplot

```{r,scatter1,fig.height=9, fig.width=9}

res <- mres1$enrichment_result

sig <- subset(res,p.adjustMANOVA < 0.01)

plot(res$s.young,res$s.old,bty="none",
  xlim=c(-1,1), ylim=c(-1,1),
  xlab="young",ylab="old",
  main="enrichment score, effect of Mst-1 overexpression") 

abline(v=0,lty=2)
abline(h=0,lty=2)

points(sig$s.young, sig$s.old, pch=19,col="black")

message("downregulated gene sets")

dn <- subset(sig,s.young < -0.5 & s.old < -0.5)$set
dn <- unname(sapply(X=dn,function(x) {substring(x, first=1, last=50) } ))
dn

message("upregulated gene sets")

up <- subset(sig,s.young > 0.5 & s.old > 0.5)$set
up <- unname(sapply(X=up,function(x) {substring(x, first=1, last=45) } ))
up

points(-0.75,-0.75,cex=30,col="red")

pos <- data.frame(-seq(0.35,1,0.05))
pos$x <- 0.05

text(0.35,-0.3,"DOWNREGULATED",cex=1,col="red")
text(pos[,2],pos[,1], head(dn,  nrow(pos)) , pos=4,cex=0.8)

points(0.75,0.75,cex=30,col="blue")

pos <- data.frame(seq(0.9,0.25,-0.05))
pos$x <- -1

text(-0.75,0.95,"UPREGULATED",cex=1,col="blue")
text(pos[,2],pos[,1], head(up,  nrow(pos)) , pos=4,cex=0.8)

```

## Get counts for heatmaps

Need to read these in from the kallisto output, clean up the rownames and then cleanup the
column names.

Then combine the 2 tables into a single one, then calculate RPM.

```{r,counts1}

tmp <- read.table("3col_young.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xy <- aggregate(. ~ geneid,x,sum)
rownames(xy) <- xy$geneid
xy$geneid = NULL
colnames(xy) <- gsub("D","Mst1tg",colnames(xy))
colnames(xy) <- gsub("E","wt",colnames(xy))
colnames(xy) <- paste(colnames(xy) ,"_young",sep="")
rpmy <- xy / colSums(xy) * 1000000
rpmy2 <- rpmy
rpmy <- rpmy[which(rowMeans(xy) > 10),]

tmp <- read.table("3col_old.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xo <- aggregate(. ~ geneid,x,sum)
rownames(xo) <- xo$geneid
xo$geneid = NULL
xo <-  xo[,grep("Gal",colnames(xo),invert=TRUE)] 
colnames(xo) <- sapply(strsplit(colnames(xo),"\\."),"[[",1)
colnames(xo) <- paste(colnames(xo) ,"_old",sep="")
rpmo <- xo / colSums(xo) * 1000000
rpmo2 <- rpmo
rpmo <- rpmo[which(rowMeans(xo) > 10),]

xx <- cbind(xy,xo)
rpm <- xx / colSums(xx) * 1000000 
rpm2 <- rpm
head(rpm)
dim(rpm)
rpm <- rpm[which(rowMeans(xx) > 10),]
dim(rpm)

```

## Mito encoded genes

```{r,mitobar1}

mty <- rpmy2[grep(" mt-",rownames(rpmy2)),]
mty <- mty[grep("mt-T",rownames(mty),invert=TRUE),]
mtywt <- mty[,grep("wt",colnames(mty))]
mtytg <- mty[,grep("tg",colnames(mty))]

mto <- rpmo2[grep(" mt-",rownames(rpmo2)),]
mto <- mto[grep("mt-T",rownames(mto),invert=TRUE),]
mtowt <- mto[,grep("wt",colnames(mto))]
mtotg <- mto[,grep("Tg",colnames(mto))]

myboxplot <- function(GNAME) {

dat <- list("young wt"=as.numeric(mtywt[GNAME,]),
    "young tg"=as.numeric(mtytg[GNAME,]) ,
    "old wt"=as.numeric(mtowt[GNAME,]),
    "old tg"=as.numeric(mtotg[GNAME,]) )

boxplot(dat, main=GNAME, col="white", 
  log="y", ylab="RPM (log scale)")

beeswarm(dat,add=TRUE,pch=19,cex=1.2)
}

GNAMES <- intersect( rownames(mto) , rownames(mty) )

myboxplot(GNAMES[1])

par(mfrow=c(2,2))

sapply(GNAMES,myboxplot)

```

Try with ggplot.

```{r,mitobar2}

mty <- rpmy2[grep(" mt-",rownames(rpmy2)),]
mtywt <- mty[,grep("wt",colnames(mty))]
mtywtl <- melt(mtywt)
mtywtl$gene <- rep(rownames(mtywt),ncol(mtywt))
mtywtl <- mtywtl[,c(3,1,2)]
colnames(mtywtl) <- c("gene","sample","quant")
mtywtl$sample <- "wt"

mtytg <- mty[,grep("tg",colnames(mty))]
mtytgl <- melt(mtytg)
mtytgl$gene <- rep(rownames(mtytg),ncol(mtytg))
mtytgl <- mtytgl[,c(3,1,2)]
colnames(mtytgl) <- c("gene","sample","quant")
mtytgl$sample <- "tg"

mtl <- rbind(mtywtl, mtytgl)

ggplot(mtl, aes(x = factor(gene), y = quant, fill = factor(sample))) + 
  geom_boxplot() 

mtl <- mtl[grep("mt-T",mtl$gene,invert=TRUE),]

ggplot(mtl, aes(x = factor(gene), y = quant, fill = factor(sample))) +
  geom_boxplot()

```

Need to show the RQ only.

```{r,mitobar3, fig.height=7, fig.width=5}

mty <- rpmy2[grep(" mt-",rownames(rpmy2)),]
mty <- mty[grep("mt-T",rownames(mty),invert=TRUE),]
mtywt <- mty[,grep("wt",colnames(mty))]
mtywt_means <- rowMeans(mtywt)
mtytg <- mty[,grep("tg",colnames(mty))]
mtytg_means <- rowMeans(mtytg)
mtywt <- mtywt / mtywt_means
mtytg <- mtytg / mtywt_means

mtywtl <- melt(mtywt)
mtywtl$gene <- rep(rownames(mtywt),ncol(mtywt))
mtywtl <- mtywtl[,c(3,1,2)]
colnames(mtywtl) <- c("gene","sample","quant")
mtywtl$sample <- "wt"

mtytgl <- melt(mtytg)
mtytgl$gene <- rep(rownames(mtytg),ncol(mtytg))
mtytgl <- mtytgl[,c(3,1,2)]
colnames(mtytgl) <- c("gene","sample","quant")
mtytgl$sample <- "tg"

mtl <- rbind(mtywtl, mtytgl)

mtl$gene <- sapply(strsplit(mtl$gene , " "),"[[",2)

ggplot(mtl, aes(x = factor(gene), y = quant, fill = factor(sample))) +
  coord_flip() +
  geom_boxplot() +
  ggtitle("young") +
  theme_bw() +
  theme(text = element_text(size = 20))  

mto <- rpmo2[grep(" mt-",rownames(rpmo2)),]
mto <- mto[grep("mt-T",rownames(mto),invert=TRUE),]
mtowt <- mto[,grep("wt",colnames(mto))]
mtowt_means <- rowMeans(mtowt)
mtotg <- mto[,grep("Tg",colnames(mto))]
mtotg_means <- rowMeans(mtotg)
mtowt <- mtowt / mtowt_means
mtotg <- mtotg / mtowt_means

mtowtl <- melt(mtowt)
mtowtl$gene <- rep(rownames(mtowt),ncol(mtowt))
mtowtl <- mtowtl[,c(3,1,2)]
colnames(mtowtl) <- c("gene","sample","quant")
mtowtl$sample <- "wt"

mtotgl <- melt(mtotg)
mtotgl$gene <- rep(rownames(mtotg),ncol(mtotg))
mtotgl <- mtotgl[,c(3,1,2)]
colnames(mtotgl) <- c("gene","sample","quant")
mtotgl$sample <- "tg"

mtl <- rbind(mtowtl, mtotgl)

mtl$gene <- sapply(strsplit(mtl$gene , " "),"[[",2)

ggplot(mtl, aes(x = factor(gene), y = quant, fill = factor(sample))) +
  coord_flip() +
  geom_boxplot() +
  ggtitle("old") +
  theme_bw() +
  theme(text = element_text(size = 20))

```

## Heatmaps

Here, I'm going to create heatmaps for some particular gene sets.

* Respiratory electron transport

* Collagen biosynthesis

* mtDNA repair

* Mitochondrial RNA binding

They may look a bit messy because combining separate experiments can be problematic.

```{r,heatmap1}

colfunc <- colorRampPalette(c("blue", "white", "red"))

rpm_genes <- sapply(strsplit(rownames(rpm)," "),"[[",2)

SETS <- unique( c( names(genesets)[grep("RESPIRATORY",names(genesets))],
    names(genesets)[grep("COLLAGEN",names(genesets))], 
    names(genesets)[grep("REPAIR",names(genesets))],
    names(genesets)[grep("MITOCHONDRIAL",names(genesets))] ) )

myheatmap <- function(SET) {
    message(SET)
    geneset_members <- genesets[[which(names(genesets)==SET)]]

    if ( length(which(rpm_genes %in% geneset_members)) > 10 ) {

        rpm_subset <- rpm[which(rpm_genes %in% geneset_members),]

        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row", 
            col=colfunc(25) , dendrogram="none", margins = c(10,15))
        mtext(SET)
    }
}

sapply(SETS,myheatmap)

```

They may look a bit messy because combining separate experiments can be problematic.

```{r,heatmap2,fig.height=7,fig.width=7}

rpmy_genes <- sapply(strsplit(rownames(rpmy)," "),"[[",2)
rpmo_genes <- sapply(strsplit(rownames(rpmo)," "),"[[",2)

myheatmap2 <- function(SET) {
    message(SET)
    geneset_members <- genesets[[which(names(genesets)==SET)]]
    if ( length(which(rpmy_genes %in% geneset_members)) > 10 ) {
        rpm_subset <- rpmy[which(rpmy_genes %in% geneset_members),]
        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row",
            col=colfunc(25) , dendrogram="none", margins = c(10,18) )
        mtext(SET)
    }
    if ( length(which(rpmo_genes %in% geneset_members)) > 10 ) {
        rpm_subset <- rpmo[which(rpmo_genes %in% geneset_members),]
        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row",
            col=colfunc(25) , dendrogram="none", margins = c(10,18))
        mtext(SET)
    }
}

sapply(SETS,myheatmap2)


```

## Session information

```{r,sessioninfo}

sessionInfo()

```
