---
title: "Reanalysis of the effect of Mst-1 overexpression in young and old mice"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/mst1_reanalysis

# Introduction

```{r,packages}

suppressPackageStartupMessages({
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("mitch")
    library("eulerr")
    library("kableExtra")
    library("beeswarm")
})

```

## Import DE data


```{r,importdata}

you <- read.table("Mst1TG_young_deseq.tsv",header=TRUE,sep="\t")

old <- read.table("Mst1TG_old_deseq.tsv",header=TRUE,sep="\t")

you_up <- rownames(subset(you,log2FoldChange > 0 & padj < 0.05))
you_dn <- rownames(subset(you,log2FoldChange < 0 & padj < 0.05))

old_up <- rownames(subset(old,log2FoldChange > 0 & padj < 0.05))
old_dn <- rownames(subset(old,log2FoldChange < 0 & padj < 0.05))

```


## Euler diagram

```{r,euler1}

par(mar=c(2,2,2,2))

v0 <- list("young up"=you_up, 
  "young dn"=you_dn,
  "old up"=old_up,
  "old dn"=old_dn)

plot(euler(v0),quantities = TRUE, edges = "gray", main="effect of Mst-1 overexpression")

```

## Mitch

First fetch genesets.

```{r,geneset1}

if  (!file.exists("mouse_msigdb_reactome_2022-02-16.gmt")) {
  download.file("http://ziemann-lab.net/public/msigdb_mouse/mouse_msigdb_reactome_2022-02-16.gmt",
    destfile="mouse_msigdb_reactome_2022-02-16.gmt")
}
genesets <- gmt_import("mouse_msigdb_reactome_2022-02-16.gmt")
names(genesets) <- gsub("REACTOME_","",names(genesets))
names(genesets) <- gsub("_"," ",names(genesets))

```

Now set up the gene table `gt`.

```{r,gt}

gt <- as.data.frame(unique(rownames(old) ,rownames(you)))
gt$gene <- sapply(strsplit(gt[,1]," "),"[[",2)

```

Now execute mitch analysis.

```{r,mitch1}

m1 <- mitch_import(list("young"=you,"old"=old), DEtype="deseq2",geneTable=gt)

mres1 <- mitch_calc(m1, genesets, priority="effect")

head(mres1$enrichment_result,20) %>% 
  kbl(caption = "Top gene pathway differences caused by Mst-1 overexpression") %>%
  kable_paper("hover", full_width = F)

```

Scatterplot

```{r,scatter1,fig.height=9, fig.width=9}

res <- mres1$enrichment_result

sig <- subset(res,p.adjustMANOVA < 0.01)

plot(res$s.young,res$s.old,bty="none",
  xlim=c(-1,1), ylim=c(-1,1),
  xlab="young",ylab="old",
  main="enrichment score, effect of Mst-1 overexpression") 

abline(v=0,lty=2)
abline(h=0,lty=2)

points(sig$s.young, sig$s.old, pch=19,col="black")

message("downregulated gene sets")

dn <- subset(sig,s.young < -0.5 & s.old < -0.5)$set
dn <- unname(sapply(X=dn,function(x) {substring(x, first=1, last=50) } ))
dn

message("upregulated gene sets")

up <- subset(sig,s.young > 0.5 & s.old > 0.5)$set
up <- unname(sapply(X=up,function(x) {substring(x, first=1, last=45) } ))
up

points(-0.75,-0.75,cex=30,col="red")

pos <- data.frame(-seq(0.35,1,0.05))
pos$x <- 0.05

text(0.35,-0.3,"DOWNREGULATED",cex=1,col="red")
text(pos[,2],pos[,1], head(dn,  nrow(pos)) , pos=4,cex=0.8)

points(0.75,0.75,cex=30,col="blue")

pos <- data.frame(seq(0.9,0.25,-0.05))
pos$x <- -1

text(-0.75,0.95,"UPREGULATED",cex=1,col="blue")
text(pos[,2],pos[,1], head(up,  nrow(pos)) , pos=4,cex=0.8)

```

## Get counts for heatmaps

Need to read these in from the kallisto output, clean up the rownames and then cleanup the
column names.

Then combine the 2 tables into a single one, then calculate RPM.

```{r,counts1}

tmp <- read.table("3col_young.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xy <- aggregate(. ~ geneid,x,sum)
rownames(xy) <- xy$geneid
xy$geneid = NULL
colnames(xy) <- gsub("D","Mst1tg",colnames(xy))
colnames(xy) <- gsub("E","wt",colnames(xy))
colnames(xy) <- paste(colnames(xy) ,"_young",sep="")
rpmy <- xy / colSums(xy) * 1000000
rpmy <- rpmy[which(rowMeans(xy) > 10),]

tmp <- read.table("3col_old.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xo <- aggregate(. ~ geneid,x,sum)
rownames(xo) <- xo$geneid
xo$geneid = NULL
xo <-  xo[,grep("Gal",colnames(xo),invert=TRUE)] 
colnames(xo) <- sapply(strsplit(colnames(xo),"\\."),"[[",1)
colnames(xo) <- paste(colnames(xo) ,"_old",sep="")
rpmo <- xo / colSums(xo) * 1000000
rpmo <- rpmo[which(rowMeans(xo) > 10),]

xx <- cbind(xy,xo)
rpm <- xx / colSums(xx) * 1000000 
head(rpm)
dim(rpm)
rpm <- rpm[which(rowMeans(xx) > 10),]
dim(rpm)

```

## Heatmaps

Here, I'm going to create heatmaps for some particular gene sets.

* Respiratory electron transport

* Collagen biosynthesis

* mtDNA repair

* Mitochondrial RNA binding

They may look a bit messy because combining separate experiments can be problematic.

```{r,heatmap1}

colfunc <- colorRampPalette(c("blue", "white", "red"))

rpm_genes <- sapply(strsplit(rownames(rpm)," "),"[[",2)

SETS <- unique( c( names(genesets)[grep("RESPIRATORY",names(genesets))],
    names(genesets)[grep("COLLAGEN",names(genesets))], 
    names(genesets)[grep("REPAIR",names(genesets))],
    names(genesets)[grep("MITOCHONDRIAL",names(genesets))] ) )

myheatmap <- function(SET) {
    message(SET)
    geneset_members <- genesets[[which(names(genesets)==SET)]]

    if ( length(which(rpm_genes %in% geneset_members)) > 10 ) {

        rpm_subset <- rpm[which(rpm_genes %in% geneset_members),]

        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row", 
            col=colfunc(25) , dendrogram="none", margins = c(10,15))
        mtext(SET)
    }
}

sapply(SETS,myheatmap)

```

They may look a bit messy because combining separate experiments can be problematic.

```{r,heatmap2,fig.height=7,fig.width=7}

rpmy_genes <- sapply(strsplit(rownames(rpmy)," "),"[[",2)
rpmo_genes <- sapply(strsplit(rownames(rpmo)," "),"[[",2)

myheatmap2 <- function(SET) {
    message(SET)
    geneset_members <- genesets[[which(names(genesets)==SET)]]
    if ( length(which(rpmy_genes %in% geneset_members)) > 10 ) {
        rpm_subset <- rpmy[which(rpmy_genes %in% geneset_members),]
        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row",
            col=colfunc(25) , dendrogram="none", margins = c(10,18) )
        mtext(SET)
    }
    if ( length(which(rpmo_genes %in% geneset_members)) > 10 ) {
        rpm_subset <- rpmo[which(rpmo_genes %in% geneset_members),]
        gnames <- sapply(strsplit(rownames(rpm_subset)," "),"[[",2)
        gnames[which(duplicated(gnames) )] <- paste( gnames[which(duplicated(gnames) )] , "2" )
        rownames(rpm_subset) <- gnames
        heatmap.2(as.matrix(rpm_subset), trace="none", scale="row",
            col=colfunc(25) , dendrogram="none", margins = c(10,18))
        mtext(SET)
    }
}

sapply(SETS,myheatmap2)


```

## Session information

```{r,sessioninfo}

sessionInfo()

```
